A new tidy data structure to support exploration and modeling of temporal data

### 2019-08-17 tsibble.Rmd
done with revision on vis

The **tidyverse** largely formalizes exploratory data analysis. Many in the R community have adopted the **tidyverse** way of thinking and extended it to broader domains, such as simple features for spatial data in the **sf** package [@RJ-2018-009] and missing value handling in the **naniar** package [@tierney-naniar-2018]. This paper and the associated **tsibble** R package extends the tidy way of thinking to temporal data.

@@

What subjects/entities to be observed over time leads to the second component of a tsibble--key. Key can consist of empty, single, or multiple variables identifying units measured along the way. When only a single observational unit is present in the table, key can be left with no variables to be specified. However, when multiple units exist in the data, key should be supplied by identifying variables to sufficiently define the units. In longitudinal data, key can be thought of as "panel" (such as in the Stata) but constrained to a single variable. In tsibble, key allows for multiple variables of nesting, crossing, or union relations [@wilkinson2006grammar] to be useful for forecasting reconciliation [@fpp; @R-hts] and other advanced models. For example, Table \@ref(tab:tb-sub) describes the number of tuberculosis cases for each gender across the countries every year. This suggests that the key comprises at least columns `gender` and `country`. Since `country` is nested within `continent`, it is a free variable to be included in the key specification.

@@

The other key verb is `index_by()`, the counterpart of `group_by()` in **dplyr**, grouping and partitioning the index only. It is most useful to be used in conjunction with `summarize()`, and thus aggregations to upper-level time resolutions will take place. This combination automatically produces new index and interval, and helps to regularize data of irregular interval.

Besides new verbs, the **dplyr** vocabulary has been adapted and expanded to facilitate temporal transformations. The **dplyr** suite showcases the general-purpose verbs for effectively manipulating tabular data. But these verbs need handling with care due to the context switch. A perceivable difference is summarizing variables between normal data and tsibble using `summarize()`. The former will reduce to a single summary, whereas the latter will obtain the index and their corresponding summaries. Attention has been paid to warning and error handling. The principle that underpins most verbs is a tsibble in and a tsibble out, thereby striving to maintain a valid tsibble by automatically updating index and key under the hood. If the desired temporal ordering is changed by row-wise verbs (such as `arrange()` and `slice()`), a warning is broadcast. If a tsibble cannot be maintained in the output of a pipeline module (likely occurring to column-wise verbs), for example the index is removed by `select()`-ing, an error informs users of the problem and suggests alternatives. This avoids surprising users and reminds them of the time context. In general, users who are already familiar with the **tidyverse**, should have less resistance to learning the new semantics and verbs.

@@

As a tsibble is a subclassing of data frame, it integrates well with the grammar of graphics. It is easy to create and extend specialist time series plotting methods based on the tsibble structure, for example autocorrelation plots and calendar-based graphics [@wang-calendarviz-2018].

@@

>>>>

The **tidyverse** largely formalizes exploratory data analysis. Many in the R community have adopted the **tidyverse** way of thinking and extended it to broader domains, such as simple features for spatial data in the **sf** package [@RJ-2018-009] and missing value handling in the **naniar** package [@tierney-naniar-2018]. This paper and the associated **tsibble** R package [@R-tsibble] extends the tidy way of thinking to temporal data.

@@

What subjects/entities to be observed over time leads to the second component of a tsibble--key. Key can consist of empty, single, or multiple variables identifying units measured along the way. When only a single observational unit is present in the table, key can be left with no variables to be specified. However, when multiple units exist in the data, key should be supplied by identifying variables to sufficiently define the units. In longitudinal data, key can be thought of as "panel" (such as in the Stata) and is constrained to a single variable. In tsibble, key allows for multiple variables of nesting, crossing, or union relations [@wilkinson2006grammar] to be useful for forecasting reconciliation [@fpp; @R-hts] and other advanced models. For example, Table \@ref(tab:tb-sub) describes the number of tuberculosis cases for each gender across the countries every year. This suggests that the key comprises at least columns `gender` and `country`. Since `country` is nested within `continent`, it is a free variable to be included in the key specification.

@@

The other key function, an adverb `index_by()`, is the counterpart of `group_by()` in **dplyr**, grouping and partitioning the index only. It is most useful to be used in conjunction with `summarize()`, and thus aggregations to upper-level time resolutions will take place. This combination automatically produces new index and interval, and helps to regularize data of irregular interval.

Besides new verbs, the **dplyr** vocabulary has been adapted and expanded to facilitate temporal transformations. The **dplyr** suite showcases the general-purpose verbs for effectively manipulating tabular data. But these verbs need handling with care due to the context switch. A perceivable difference is summarizing variables between normal data and tsibble using `summarize()`. The former will reduce to a single summary, whereas the latter will obtain the index and their corresponding summaries. Attention has been paid to warning and error handling. The principle that underpins all the verbs is a tsibble in and a tsibble out, thereby striving to maintain a valid tsibble over the course of transformation. If the desired temporal ordering is changed by row-wise verbs (such as `arrange()` and `slice()`), a warning is broadcast. If a tsibble cannot be maintained in the output of a pipeline module (likely occurring to column-wise verbs), for example the index is removed by `select()`-ing, an error informs users of the problem and suggests alternatives. This avoids surprising users and reminds them of the time context. In general, users who are already familiar with the **tidyverse**, should have less resistance to learning the new semantics and verbs.

@@

The **ggplot2** package (as the implementation of grammar of graphics) builds a powerful graphical system to declaratively visualize data. The data underpinning of **ggplot2** is tidy data, and in turn tsibble integrates well with **ggplot2**. The integration encourages more flexible graphics for exploring temporal structures, and individual or group differences.

Line charts are universally accepted for ordered data, such as time series plots or spaghetti plots, depending on the fields. But they end up with exactly the same grammar: chronological time mapped to the horizontal axis, and the interested measurement on the vertical axis, for each unit. Many specialist plots centering around time series or longitudinal data, hence can be described and re-created by the grammar and **ggplot2**.

@@

* scalable
* human-centered
